Demographic Influences on Municipal Budget Allocation and Infrastructure Development in Toronto

### 2024-12-15 paper/paper.qmd
Grammer Fix
- Fixed grammer

---
title: "Demographic Influences on Municipal Budget Allocation and Infrastructure Development in Toronto"
subtitle: "Population density drive larger budget allocations meanwhile average household income drives construction projects"
author: 
  - Aviral Bhardwaj
thanks: "Code and data are available at: [https://github.com/Aviral-03/InfrastructureCausalModel-TO](https://github.com/Aviral-03/InfrastructureCausalModel-TO)." 
date: today
date-format: long
abstract: "This study examines the relationship between population density, average household income, and budget allocations across Toronto's 25 wards and investigates their combined influence on construction activity. Using data from the city's *Capital Budget Plan*, *Ward Profiles*, and *Active Building Permits*, we analyze spending patterns in key infrastructure-related areas. Employing a causal modeling approach, our findings reveal a negative association between population density and construction activity, despite some high-density wards receiving higher budget allocations. Additionally, average household income and total budget allocations exhibit a positive correlation with the number of building permits. These results underscore potential inequities in resource distribution, highlighting the need for data-driven and equitable urban budget planning to better address community needs."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
prefer-html: true
---

@@

The paper is organized as follows: [@sec-data] outlines the data sources and methodology used in the analysis. [@sec-model] explains the Bayesian model applied to estimate the effects of demographic and economic factors, following the causal model ([@fig-causal-model]). [@sec-results] presents key findings, including trends in budget allocations, the relationship between demographic factors and budget distribution, and the impact on construction projects. [@sec-discussion] explores the implications for urban governance and budget planning in Toronto. Finally, [@sec-limitations] addresses the study’s limitations and offers recommendations for future research. Additional details, including tables, figures, and methodology, are provided in the [@sec-appendix].

@@

Ward 10 (Spadina-Fort York) received the highest budget allocations across both *Growth Related* and *State of Good Repair* categories, indicating significant investment in infrastructure development and maintenance to population-dense areas. Toronto Center (Ward 13), Don Valley East (Ward 16), and Etobicoke Center (Ward 2) also received notable budget allocations, some of these wards have higher average household incomes as well as higher population densities. The box plots reveal that budget allocations for *Growth Related* projects have increased steadily over the years, with 75% of wards receiving higher allocations in 2024 compared to 2021. In contrast, budget allocations for *State of Good Repair* and *Service Improvement and Enhancement* projects have remained relatively consistent, with few outliers receiving higher budget allocations.

@@

This analysis has several limitations that warrant consideration. As discussed in [@sec-data], the datasets used, sourced from the City of Toronto, provide insights into the city’s demographics, budget allocations, and building permits. However, they may not fully capture the complexity of Toronto's infrastructure development and resource distribution.

@@

Additionally, as noted in [@sec-appendix], there are 45 unique statuses assigned to building permits, but this study analyzed only a subset of these statuses. This narrow focus may overlook important stages in the construction process. Expanding the analysis to include more permit statuses could offer a more comprehensive understanding of construction activity.

Another limitation is the lack of detailed data on specific projects within each budget category, which restricts insights into why certain wards receive more or less funding. Access to granular project-level data could provide deeper context and explanations for funding disparities.

>>>>

---
title: "Demographic Influences on Municipal Budget Allocation and Infrastructure Development in Toronto"
subtitle: "Population density influences larger budget allocations, while average household income determines the focus of construction projects"
author: 
  - Aviral Bhardwaj
thanks: "Code and data are available at: [https://github.com/Aviral-03/InfrastructureCausalModel-TO](https://github.com/Aviral-03/InfrastructureCausalModel-TO)." 
date: today
date-format: long
abstract: "This study examines the relationship between population density, average household income, and budget allocations across Toronto's 25 wards and investigates their combined influence on construction activity. Using data from the city's *Capital Budget Plan*, *Ward Profiles*, and *Active Building Permits*, we analyze spending patterns in key infrastructure-related areas. Employing a causal modeling approach, our findings reveal a negative association between population density and construction activity, despite some high-density wards receiving higher budget allocations. Additionally, average household income and total budget allocations exhibit a positive correlation with the number of building permits. These results underscore potential inequities in resource distribution, highlighting the need for data-driven and equitable urban budget planning to better address community needs."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
prefer-html: true
---

@@

The paper is organized as follows: [@sec-data] outlines the data sources and methodology used in the analysis. [@sec-model] explains the Bayesian model applied to estimate the effects of demographic and economic factors, following the causal model ([@fig-causal-model]). [@sec-results] presents key findings, including trends in budget allocations, the relationship between demographic factors and budget distribution, and the impact on construction projects. [@sec-discussion] explores the implications for urban governance and budget planning in Toronto. Finally, [@sec-limitations] addresses the study’s limitations and suggests recommendations for future research. Additional details, including tables, figures, and methodology, are provided in the [@sec-appendix].

@@

Ward 10 (Spadina-Fort York) received the highest budget allocations across both *Growth Related* and *State of Good Repair* categories, indicating significant investment in infrastructure development and maintenance to population-dense areas. Toronto Center (Ward 13), Don Valley East (Ward 16), and Etobicoke Center (Ward 2) also received notable budget allocations, some of these wards have higher average household incomes as well as higher population densities. The box plots presents that budget allocations for *Growth Related* projects have increased steadily over the years, with 75% of wards receiving higher allocations in 2024 compared to 2021. In contrast, budget allocations for *State of Good Repair* and *Service Improvement and Enhancement* projects have remained relatively consistent, with few outliers receiving higher budget allocations.

@@

This analysis has several limitations that warrant consideration. As discussed in [@sec-data], the datasets used, sourced from the City of Toronto, presents city’s demographics, budget allocations, and building permits. However, they may not fully capture the complexity of Toronto's infrastructure development and resource distribution.

@@

Additionally, as noted in [@sec-appendix], there are 45 unique statuses assigned to building permits, but this study analyzed only a subset of these statuses. This narrow focus may overlook important stages in the construction process. Expanding the analysis to include more permit statuses could offer a more in-depth understanding of construction activity.

Another limitation is the lack of detailed data on specific projects within each budget category, which restricts the paper's analysis into why certain wards receive more or less funding. Access to granular project-level data could provide deeper context and explanations for funding disparities.
